Bayesian multivariate skew-normal finite mixture model for analysis of infant development trajectories

### 2019-07-09 manuscript/manuscript.tex
sim 1

Our first simulation study compared the MSN mixture model to a MVN mixture model, with the primary goal being to validate our parameter estimation scheme in a setting that resembles the Nurture data. Our secondary goal was to investigate to what degree ignoring skewness in outcome components leads to poor posterior inferences. To emulate the Nurture study, we simulated $n = 1,000$ subjects from the following model
\begin{eqnarray}
	\label{eq:sim1}
	f(\mathbf{y}_i) = \sum_{k = 1}^3 \pi_{ki} f(\mathbf{y}_i | \boldsymbol\theta_k),
\end{eqnarray}
where $\mathbf{y}_i = (y_{i1},...,y_{i4})^T$ to conform to the $J = 4$ measurement occasions in the Nurture study; $\boldsymbol\theta_k$ is the set of parameters specific to cluster $k$ ($k = 1,2,3$), and $f(\mathbf{y}_i | \boldsymbol\theta_k) \stackrel{\text{d}}{=} \text{MSN}_4(\boldsymbol\zeta_{ki},\boldsymbol\alpha_k,\boldsymbol\Omega_k)$; $\boldsymbol\zeta_{ki} = (\zeta_{ki1},...,\zeta_{ki4})^T = \mathbf{X}_i \boldsymbol\beta_{k}$, $\boldsymbol\beta_k = \text{vec}(\mathbf{B}_k)$, and $\mathbf{X}_i$ and $\mathbf{B}_k$ are given by. 
\begin{eqnarray}
	\mathbf{x}_i = 
	\begin{bmatrix}
		1 & 0 & 0 & 0 & x_i & 0 & 0 & 0\\
		0 & 1 & 0 & 0 & 0 & x_i & 0 & 0\\
		0 & 0 & 1 & 0 & 0 & 0 & x_i & 0\\
		0 & 0 & 0 & 1 & 0 & 0 & 0 & x_i\\
	\end{bmatrix}, \ \text{and} \ 
	\mathbf{B}_k = 
	\begin{bmatrix}
		\beta_{k11} & \beta_{k12} \\
		\beta_{k21} & \beta_{k22} \\
		\beta_{k31} & \beta_{k32} \\
		\beta_{k41} & \beta_{k42} \\
	\end{bmatrix}. \nonumber
\end{eqnarray}
We fit a dummy indicator for each measurement occasion and a time invariant baseline covariate $x_i$ (e.g., birthweight for gestational age z-score), whose effect may vary across the $J$ measurement occasions. Thus, for subject $i$ in cluster $k$, the $J = 4$ measurements have location parameters given by
\begin{eqnarray}
	\zeta_{ki1} &=& \beta_{k11} + \beta_{k12}x_i \nonumber \\
	\zeta_{ki2} &=& \beta_{k21} + \beta_{k22}x_i \nonumber \\
	\zeta_{ki3} &=& \beta_{k31} + \beta_{k32}x_i \nonumber \\
	\zeta_{ki4} &=& \beta_{k41} + \beta_{k42}x_i. \nonumber 
\end{eqnarray}
To emulate z-score-transformed covariates present in the Nurture data such as birthweight for gestational age z-score, we sampled $x_i$ from a $\text{N}(0,1)$ density for $i = 1,...,n$.

For cluster $k$, we collected all observations $\mathbf{y}_i$ ($i = 1,...,n_k$) into the $n_k \times 4$ matrix $\mathbf{Y}_k$ and all covariates $\mathbf{X}_i$ into the $4n_k \times 8$ matrix $\mathbf{X}_k$ to utilize the matrix normal model specification detailed in Section 3, which allows for simultaneous updates of all MSN regression parameters in cluster $k$. For the multinomial regression model component of this simulation, we modeled the class labels $z_i$ as a function of an intercept and one baseline covariate, $\text{w}_{i1}$, implying that $r = 2$. We did introduce missing data into this simulation, as missing data will be dealt with in our second simulation study. As a result, the final sample size was $N = n \times J = 4,000$.

@@

In our second simulation study, we compare our online conditional imputation approach to standard Bayesian multiple imputation as defined in Gelman et al. (2013). For an argument against the use of data-discarding approaches such as available case analysis (ACA) or complete case analysis (CCA), we defer to Gelman and Hill (2006). Our goal with this simulation is to demonstrate that our proposed online conditional imputation method performs as well as, if not better than standard Bayesian multiple imputation with respect to accuracy of parameter estimates. If the performance of the two imputation methods are held equal, we argue that our online conditional imputation method is preferable due to its computational efficiency and avoidance of a burdensome and often error-prone parameter aggregation step in the analysis.

To demonstrate the advantages of our proposed imputation method, we first generate $n = 1,000$ observations from a simple three-cluster ($K = 3$) MSN mixture model with $J = 4$ repeated measurements, one main effect for time ($p = 1$), and two multinomial regression predictors ($r = 2$). We then artificially ampute the simulated data using the \texttt{ampute} function provided by the \texttt{mice} package in \texttt{R} (Van Buuren and Groothuis-Oudshoorn 2010). Amputation occurs according to the missing at random (MAR) mechanism, whereby the occurence of missing data depends only on the values of the observed data (Gelman and Hill 2006). Additionally, we perform data amputation to simulateintermittent missingness, as is consistent with the Nurture data. To this end, we specify that each of the $4$ repeated measures for a given individual are equally likely to be missing.  To further approximate the Nurture data, we set the frequency of missing data at each measurement occasion to be 30\%. We fit our proposed MSN finite mixture model to the amputed data with an implementation of both our proposed online conditional imputation step and standard Bayesian multiple imputation as defined by Gelman et al. (2013).

>>>>

Our first simulation study compared the MSN mixture model to a MVN mixture model, with the primary goal being to validate our parameter estimation scheme in a setting that resembles the Nurture data. Our secondary goal was to investigate to what degree ignoring skewness in outcome components leads to poor posterior inferences. To emulate the Nurture study, we simulated $n = 1,000$ subjects from the following model
\begin{eqnarray}
	\label{eq:sim1}
	f(\mathbf{y}_i) = \sum_{k = 1}^3 \pi_{ki} f(\mathbf{y}_i | \boldsymbol\theta_k),
\end{eqnarray}
where $\mathbf{y}_i = (y_{i1},...,y_{i4})^T$ to conform to the $J = 4$ measurement occasions in the Nurture study; $\boldsymbol\theta_k$ is the set of parameters specific to cluster $k$ ($k = 1,2,3$), and $f(\mathbf{y}_i | \boldsymbol\theta_k) \stackrel{\text{d}}{=} \text{MSN}_4(\boldsymbol\zeta_{ki},\boldsymbol\alpha_k,\boldsymbol\Omega_k)$; $\boldsymbol\zeta_{ki} = (\zeta_{ki1},...,\zeta_{ki4})^T = \mathbf{X}_i \boldsymbol\beta_{k}$, $\boldsymbol\beta_k = \text{vec}(\mathbf{B}_k)$, and $\mathbf{X}_i$ and $\mathbf{B}_k$ are given by. 
\begin{eqnarray}
	\mathbf{X}_i = 
	\begin{bmatrix}
		1 & 0 & 0 & 0 & x_i & 0 & 0 & 0\\
		0 & 1 & 0 & 0 & 0 & x_i & 0 & 0\\
		0 & 0 & 1 & 0 & 0 & 0 & x_i & 0\\
		0 & 0 & 0 & 1 & 0 & 0 & 0 & x_i\\
	\end{bmatrix}, \ \text{and} \ 
	\mathbf{B}_k = 
	\begin{bmatrix}
		\beta_{k11} & \beta_{k12} \\
		\beta_{k21} & \beta_{k22} \\
		\beta_{k31} & \beta_{k32} \\
		\beta_{k41} & \beta_{k42} \\
	\end{bmatrix}. \nonumber
\end{eqnarray}
We fit a dummy indicator for each measurement occasion and a time invariant baseline covariate $x_i$ (e.g., birthweight for gestational age z-score), whose effect may vary across the $J$ measurement occasions. Thus, for subject $i$ in cluster $k$, the $J = 4$ measurements have location parameters given by
\begin{eqnarray}
	\zeta_{ki1} &=& \beta_{k11} + \beta_{k12}x_i \nonumber \\
	\zeta_{ki2} &=& \beta_{k21} + \beta_{k22}x_i \nonumber \\
	\zeta_{ki3} &=& \beta_{k31} + \beta_{k32}x_i \nonumber \\
	\zeta_{ki4} &=& \beta_{k41} + \beta_{k42}x_i. \nonumber 
\end{eqnarray}
To emulate z-score-transformed covariates present in the Nurture data such as birthweight for gestational age z-score, we sampled $x_i$ from a $\text{N}(0,1)$ density for $i = 1,...,n$.

For cluster $k$, we collected all observations $\mathbf{y}_i$ ($i = 1,...,n_k$) into the $n_k \times 4$ matrix $\mathbf{Y}_k$ and all covariates $\mathbf{X}_i$ into the $4n_k \times 8$ matrix $\mathbf{X}_k$ to utilize the matrix normal model specification detailed in Section 3, which allows for simultaneous updates of all MSN regression parameters in cluster $k$. For the multinomial regression model component of this simulation, we modeled the class labels $z_i$ as a function of an intercept and one baseline covariate, $\text{w}_{i1}$, implying that $r = 2$. We did introduce missing data into this simulation, as missing data will be dealt with in the second simulation study. As a result, the final sample size was $N = n \times J = 4,000$.

@@

In the second simulation study, we compare our online conditional imputation approach to Bayesian multiple imputation, in which M (large) imputed data sets are generated, MCMC is performed on each imputed data set, and the data are pooled for final posterior inference. Details of this approach are nicely summarized in (Zhou and Reiter, 2010). Our goal here is to demonstrate that our proposed online conditional imputation method performs comparable to Bayesian multiple imputation with respect to accuracy of parameter estimates. If the performance of the two methods are qualitatively alike, we might argue that the online conditional imputation method is preferable due to its computational efficiency. Additionally, we compare both methods to the default analysis based on available cases, despite extensive literature cautioning against this approach (Gelman and Hill, 2006)

To demonstrate the advantages of our proposed imputation method, we first generate $n = 1,000$ observations from a simple three-cluster ($K = 3$) MSN mixture model with $J = 4$ repeated measurements, one main effect for time ($p = 1$), and two multinomial regression predictors ($r = 2$). We then artificially ampute the simulated data using the \texttt{ampute} function provided by the \texttt{mice} package in \texttt{R} (Van Buuren and Groothuis-Oudshoorn 2010) according to the missing at random (MAR) mechanism, whereby the occurence of missing data depends only on the values of the observed data (Gelman and Hill 2006). We specified that each of the $4$ repeated measures for a given individual were equally likely to be missing.  To further approximate the Nurture data, we set the frequency of missing data at each measurement occasion to be 30\%. We then fit our proposed MSN finite mixture model to the amputed data using both online conditional imputation and Bayesian multiple imputation as defined by (Zhou and Reiter, 2010).
